Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-14 doc/invt.tex
invt.tex: completing the LJ results

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
Ideally,
the optimal initial updating magnitude
can be set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}
can be cast as special cases of bandpass updating schemes.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated at a
constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  &+
  \sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  \notag \\
  &= \frac{ M\bigl( q(T) \bigr) } { T } \int_0^{q(T)} M(Q') \, dQ'
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(Q) \, dQ
  =
  \frac{ 2 \, C } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(t = 0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \notag %\label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible, then
%
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.
}

@@

\item \label{step:prerun}
Run a preliminary adaptive FDS simulation
%using the updating scheme
with a constant updating magnitude, $a_0$.

@@

Our test system is a Lennard-Jones fluid of $N=100$ particles,
and we wish to compute the PMF along the distance of two special particles, $r$.
%
The two special particles are mutually non-interacting
but they interact normally with other particles,\footnote{they
have placed on an axis parallel to a side of the cubic box}
and the intrinsic distribution, $p^*$, is
the cavity distribution function, $y(r)$\cite{hansen}.
%
We will assume reduced units in this calculation.
%
The reference bias potential was pre-calculated from a long simulation.

@@

As shown in Table 1,
both the inverse-time and optimal schedule reduced
the error significantly,
and the errors roughly matched the predicted values
calculated from the precomputed $\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian scheme appeared to be superior
to the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
In either case, there was an optimal width of Gaussian
(which was $\sigma = 0.37$ for $\varrho^* = 0.1$
or $0.07$ for $\varrho^* = 0.8$),
an overly large width was counterproductive.
%
In Figure \ref{fig:lj_xerr},
we plotted the error components against the eigenmode index
for the Gaussian updating scheme of width $\sigma = 0.2$.
%
While the Gaussian updating scheme effectively suppressed
the initial errors for short-range (large $k$) modes,
it also slowed down the reduction of errors of these modes.
%
As a result, the final errors of the long-range (small $k$) modes
were larger than those from the single-bin scheme.
%
For the Gaussian updating scheme
the inverse-time schedule tended to reduce the updating magnitude too rapid
leaving a hump on the error profile for the middle-range modes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.

\begin{table}[h]\footnotesize
  \caption{\label{tab:error_LJ}
    Error of the LJ system.
    The numbers have been averaged over multiple runs,
    and the numbers in parentheses are predicted values;
    $\Err_\mathrm{init.}$
    denotes the error
    at the end of the equilibration phase with a constant magnitude;
    $\Err_\mathrm{final}$
    denotes the error at the end of the production phase under
    either the inverse-time ($1/t$) or optimal schedule (opt.).
    \note{The reference values can be found on the second line of \texttt{verr.log}.}
  }
  \setlength{\tabcolsep}{2pt}
  \renewcommand\arraystretch{1.4}
  \begin{tabular} { l c c c c }
    \hline
    Scheme & Sched. &
    $\Err_\mathrm{init.}$ &
    $\Err_\mathrm{final}$
    \\
    \hline
    \multicolumn{4}{c}{$\varrho^* = 0.1$, $500$ bins} \\
    \hline
    single-bin & $1/t$
    & $5.7(5.7)\times10^{-2}$
    & $9.1(11.4)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.1$ & $1/t$
    & $7.7(8.9)\times10^{-3}$
    & $6.8(8.2)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.1$ & opt.
    & $7.6(8.9)\times10^{-3}$
    & $3.2(3.9)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.2$ & $1/t$
    & $6.5(7.8)\times10^{-3}$
    & $4.0(4.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.2$ & opt.
    & $7.0(7.8)\times10^{-3}$
    & $2.5(2.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.5$ & $1/t$
    & $5.5(6.6)\times10^{-3}$
    & $3.2(3.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.5$ & opt.
    & $5.8(6.6)\times10^{-3}$
    & $2.6(3.0)\times10^{-5}$
    \\
    %\hline
    $K=20$ & $1/t$
    & $8.1(8.9)\times10^{-3}$
    & $1.6(1.8)\times10^{-5}$
    \\
    \hline
    \multicolumn{4}{c}{$\varrho^* = 0.8$, $250$ bins} \\
    \hline
    single-bin & $1/t$
    & $1.3(1.4)$
    & $2.7(2.7)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.1$ & $1/t$
    & $1.3(1.3)$
    & $2.9(2.8)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.1$ & opt.
    & $1.2(1.3)$
    & $2.8(2.8)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.2$ & $1/t$
    & $1.3(1.2)$
    & $3.4(3.2)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.2$ & opt.
    & $1.2(1.2)$
    & $3.1(2.9)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.5$ & $1/t$
    & $0.9(0.9)$
    & $9.8(8.0)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.5$ & opt.
    & $1.0(0.9)$
    & $5.8(5.4)\times10^{-3}$
    \\
    %\hline
    $K=20$ & $1/t$
    & $1.3(1.3)$
    & $3.0(2.7)\times10^{-3}$
    \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/lj_xerr.pdf}
  }
  \caption{
    \label{fig:lj_xerr}
    Error components for the FDS simulation
    using the Gaussian updating scheme with $\sigma = 0.2$.
    The optimal schedule (opt.) is more effective
    in reducing the errors of middle-range modes
    than the inverse-time ($1/t$) schedule.
    The points are joined as a guide to the eyes.
  }
\end{center}
\end{figure}

@@

The optimal schedule depends on the updating window function,
which is limited to a single-bin in the WL case,
but can be extended to several neighboring bins
taking a Gaussian shape in the metadynamics case.
%
More generally,
each adaptive FDS method is associated with
an updating scheme
that can be mathematically characterized
by an updating matrix.
%
Each column of the matrix gives the normalized
window function for the bin corresponding to the column.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
For the single-bin (WL) scheme,
the optimal schedule from our method
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule is not given by a simple closed form,
but is implicitly given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.
%
The initial updating magnitude is always
half of the previous equilibrium value.
% the value used during equilibration.

@@

Although the bandpass updating schemes
appeared to outperform similar Gaussian schemes
in our model study,
they are global in nature
and hence more expensive.
%
However, the performance gain
is relatively small for short simulations
with a local sampling process such as MD.
%
\note{
One way to improve the Gaussian updating schemes
is to use it with a hard cutoff of eigenmodes
(as we did for Fig. \ref{fig:optacmp}).
%
Then the optimal schedule
would be similar to the inverse-time one,
and the error, with the short-wavelength eigenmodes filtered,
would be close to that of the bandpass scheme.
}

@@

The optimality of the inverse-time schedule can be considered
as a natural consequence of the stabilization of a runtime average
as the sample size increases\cite{
  marsili2006, barducci2008}.
%
We first note that a runtime average
\begin{equation}
  X(t) = \frac{x(1) + \cdots + x(t)}{t}
  .
  \label{eq:avx}
\end{equation}
can be written as a recurrence relation
\begin{equation}
  X(t) = X(t - 1) + \frac{x(t) - X(t-1)}{t}
  .
  \label{eq:avx_recur}
\end{equation}
%
Here $x(t) - X(t-1)$ serves as an independent correction
from data point at time $t$ to the previous average value,
and this correction is absorbed into the average
with a weight of $1/t$,
which is the inverse sample size.

We can use Eq. \eqref{eq:avx_recur} to incrementally
improve the estimated
bias potential given by Eq. \eqref{eq:vcorr_equil}.
%
Assuming the correction, $\hat u_i$, is done at every step,
then
%
\begin{equation*}
  \hat u_i(t+1) = u_i(1) +
  \ln \frac{ H_i(t) } { p_i }
  \approx
  u_i(1) +
  \frac{ H_i(t) } { p_i } - 1
  .
\end{equation*}
%
Here the normalized histogram, $H_i(t)$,
is a runtime average of the instantaneous histogram, $h_i(t)$,
%
\begin{equation}
  H_i(t) = \frac 1 t \sum_{\tau = 1}^t h_i(\tau),
  \label{eq:hav_def}
\end{equation}
%
and we have assumed that it
is sufficiently precise in the second step.
%
Applying Eq. \eqref{eq:avx_recur} to the histogram yields
%
\begin{align}
  H_i(t)
  &=
  H_i(t-1) + \frac{ h_i(t) - H_i(t-1) } { t }
  .
  \label{eq:avH_recur}
\end{align}
%
Thus,
%Since $H_i(t)$ is the average of the instantaneous histogram, $h_i(t)$,
%we may apply Eq. \eqref{eq:av_recur}, and
%
\begin{align}
  \hat u_i(t+1)
  &=
  \hat u_i(t)
  +
  \frac{ H_i(t) - H_i(t-1) }{ p_i }
  \notag
  \\
  &=
  \hat u_i(t)
  +
  \frac{1}{t} \left[
    \frac{ h_i(t) }{ p_i }
   -\frac{ H_i(t-1) }{ p_i }
  \right]
  .
  \label{eq:vhat_update}
\end{align}
%
In the long time limit, we may assume that $H_i(t-1) \to p_i$
and the second term in the square bracket
becomes a constant shift that can be safely dropped.
%
Then Eq. \eqref{eq:vhat_update} is identical to
the single-bin updating scheme, Eq. \eqref{eq:wl_update},
under the inverse-time schedule, Eq. \eqref{eq:alpha_invt}.
%
In other words, the bias potential obtained from the latter approach
represents a runtime average.
%Thus, the bias potential from a WL simulation
%(which typically uses the single-bin updating scheme)
%under the inverse-time schedule
%would reach a similar precision
%as an ideal multicanonical simulation
%with fixed and nearly perfect bias potential.
%the precision at long times.
%%
%Thus, it is unnecessary to fix the bias potential
%after an adaptive FDS simulation has reached
%sufficiently convergent just for better precision
%of the final PMF.

>>>>

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}
can be cast as special cases of bandpass updating schemes.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated at a
constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  &+
  \sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  \notag \\
  &= \frac{ M\bigl( q(T) \bigr) } { T } \int_0^{q(T)} M(Q') \, dQ'
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(Q) \, dQ
  =
  \frac{ 2 \, C } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible, then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.
}

@@

\item \label{step:prerun}
Run a preliminary adaptive FDS simulation
%using the updating scheme
with constant updating magnitude, $a_0$.

@@

Our test system is a Lennard-Jones (LJ) fluid of $N=100$ particles,
and we wish to compute the PMF along the distance of two special particles, $r$.
%
The two special particles are mutually non-interacting
but they interact normally with other particles,\footnote{they
have placed on an axis parallel to a side of the cubic box}
and the intrinsic distribution, $p^*$, is
the cavity distribution function, $y(r)$\cite{hansen}.
%
We will assume reduced units in this calculation.
%
The reference bias potential was pre-calculated from a long simulation.

@@

As shown in Table 1,
both the inverse-time and optimal schedule reduced
the error significantly,
and the errors roughly matched the predicted values
calculated from the precomputed $\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian scheme appeared to be superior
to the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
In either case, there was an optimal width of Gaussian
(which was $\sigma = 0.37$ for $\varrho^* = 0.1$
or $0.07$ for $\varrho^* = 0.8$ in this case),
an overly large width could be counterproductive.

In Fig. \ref{fig:lj_xerr},
we show the error components against the eigenmode index
for the Gaussian updating scheme of width $\sigma = 0.2$.
%
While the Gaussian updating scheme effectively suppressed
the initial errors for short-range (large $k$) modes,
it also slowed down the reduction of errors of these modes.
%
As a result, the final errors of the long-range (small $k$) modes
were larger than those from the single-bin scheme.
%
For the Gaussian updating scheme
the inverse-time schedule tended to reduce the updating magnitude too rapid
leaving a hump on the error profile for the middle-range modes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
It is also possible to roughly
recover the single-bin result by using the correction formula
based on the histogram, Eq. \eqref{eq:UHcorr}.
%
This correction usually improves the accuracy of the small $k$ modes
but not necessarily for the large $k$ modes.

In Fig. \ref{fig:lj_alpha}. 
we show the optimal schedules in a few cases.
%
Generally, the optimal schedule
for the $\varrho^* = 0.8$ case
was closer to the inverse-time schedule
than that for the $\varrho^* = 0.1$ case.
%
This was probably because in the former case
the first mode dominates the total error,
$\Gamma_1 \approx \Gamma$,
and the schedule was roughly optimized for this mode,
resulting in
$\alpha(t) \approx 1/[\lambda_1 (t + t_0)] \approx 1/(t+t_0)$
as $\lambda_1 \approx 1$
[especially when width of the Gaussian is small,
cf. Eq. \eqref{eq:lambda_Gaussian}].
%
Since we have chosen a long equilibration period
for preliminary run with a constant updating magnitude,
the starting magnitude was usually around $a_0/2 = 5\times10^{-5}$
[cf. Eq. \eqref{eq:half_alpha0}].
%
But if the Gaussian is too wide,
such as the $\sigma = 0.6$ case,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in nonnegligible $\epsilon_k$,
and much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, our method would fail
because the resulting bias potential would no longer be small.

\begin{table}[h]\footnotesize
  \caption{\label{tab:error_LJ}
    Error of the LJ system.
    The numbers have been averaged over multiple runs,
    and the numbers in parentheses are predicted values;
    $\Err_\mathrm{init.}$
    denotes the error
    at the end of the equilibration phase with a constant magnitude;
    $\Err_\mathrm{final}$
    denotes the error at the end of the production phase under
    either the inverse-time ($1/t$) or optimal schedule (opt.).
    \note{The reference values can be found on the second line of \texttt{verr.log}.}
  }
  \setlength{\tabcolsep}{2pt}
  \renewcommand\arraystretch{1.4}
  \begin{tabular} { l c c c c }
    \hline
    Scheme & Sched. &
    $\Err_\mathrm{init.}$ &
    $\Err_\mathrm{final}$
    \\
    \hline
    \multicolumn{4}{c}{
      $\varrho^* = 0.1$,
      $500$ bins,
      $\Gamma_1 \approx 111$,
      $\Gamma \approx 1.1\times10^3$.
    } \\
    \hline
    single-bin & $1/t$
    & $5.7(5.7)\times10^{-2}$
    & $9.1(11.4)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.1$ & $1/t$
    & $7.7(8.9)\times10^{-3}$
    & $6.8(8.2)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.1$ & opt.
    & $7.6(8.9)\times10^{-3}$
    & $3.2(3.9)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.2$ & $1/t$
    & $6.5(7.8)\times10^{-3}$
    & $4.0(4.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.2$ & opt.
    & $7.0(7.8)\times10^{-3}$
    & $2.5(2.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.5$ & $1/t$
    & $5.5(6.6)\times10^{-3}$
    & $3.2(3.8)\times10^{-5}$
    \\
    %\hline
    $\sigma=0.5$ & opt.
    & $5.8(6.6)\times10^{-3}$
    & $2.6(3.0)\times10^{-5}$
    \\
    %\hline
    $K=20$ & $1/t$
    & $8.1(8.9)\times10^{-3}$
    & $1.6(1.8)\times10^{-5}$
    \\
    \hline
    \multicolumn{4}{c}{
      $\varrho^* = 0.8$,
      $250$ bins,
      $\Gamma_1 \approx 2.0\times10^4$,
      $\Gamma \approx 2.7\times10^4$
    } \\
    \hline
    single-bin & $1/t$
    & $1.3(1.4)$
    & $2.7(2.7)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.1$ & $1/t$
    & $1.3(1.3)$
    & $2.9(2.8)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.1$ & opt.
    & $1.2(1.3)$
    & $2.8(2.8)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.2$ & $1/t$
    & $1.3(1.2)$
    & $3.4(3.2)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.2$ & opt.
    & $1.2(1.2)$
    & $3.1(2.9)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.5$ & $1/t$
    & $0.9(0.9)$
    & $9.8(8.0)\times10^{-3}$
    \\
    %\hline
    $\sigma=0.5$ & opt.
    & $1.0(0.9)$
    & $5.8(5.4)\times10^{-3}$
    \\
    %\hline
    $K=20$ & $1/t$
    & $1.3(1.3)$
    & $3.0(2.7)\times10^{-3}$
    \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/lj_xerr.pdf}
  }
  \caption{
    \label{fig:lj_xerr}
    Error components for the FDS simulation on the LJ system
    using the Gaussian updating scheme with width $\sigma = 0.2$.
    %
    The optimal schedule (opt.) is more effective
    in reducing the errors of middle-range modes
    than the inverse-time ($1/t$) schedule.
    The lines are a guide to the eyes.
  }
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/lj_alpha.pdf}
  }
  \caption{
    \label{fig:lj_alpha}
    Optimal schedule of the Gaussian updating scheme
    for the LJ system.
    %
    The starting updating magnitude was usually
    around $a_0/2 = 5\times10^{-5}$
    as given by Eq. \eqref{eq:half_alpha0}.
    But an overly wide Gaussian window ($\sigma = 0.6$)
    resulted in a much larger updating magnitude.
  }
\end{center}
\end{figure}

@@

The optimal schedule depends on the updating window function,
which can be limited to a single-bin (as often in the WL case),
or extended to several neighboring bins
taking a Gaussian shape (as often in the metadynamics case).
%
More generally,
each adaptive FDS method is associated with
an updating scheme
that can be mathematically characterized
by an updating matrix.
%
Each column of the matrix gives the normalized
window function for the bin corresponding to the column.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
For the single-bin scheme,
we confirmed that the optimal schedule
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one,
the optimal schedule is not given by a simple closed form,
but is implicitly given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.
%
%The initial updating magnitude is always
%half of the previous equilibrium value.
% the value used during equilibration.

@@

We believe that one of the main appeals
of the Gaussian updating scheme
is its ability to maintain a smooth bias potential,
which is useful for MD applications.
%
However, it requires a careful choice of the width and
a sufficiently long equilibration phase under constant updating magnitude.
%
For reasonable parameters,
one may use the histogram correction, Eq. \eqref{eq:UHcorr},
in a post-simulation analysis
to recover the possible accuracy loss
and obtain a bias potential similar to
that from the optimized single-bin scheme.
%%
%However, the performance gain
%is relatively small for short simulations
%with a local sampling process such as MD.

@@

The optimality of the inverse-time schedule can be considered
as a natural consequence of the stabilization of a runtime average
as the sample size increases\cite{
  marsili2006, barducci2008}.
%
%We first note that a runtime average
%\begin{equation}
%  X(t) = \frac{x(1) + \cdots + x(t)}{t}
%  .
%  \label{eq:avx}
%\end{equation}
%can be written as a recurrence relation
%\begin{equation}
%  X(t) = X(t - 1) + \frac{x(t) - X(t-1)}{t}
%  .
%  \label{eq:avx_recur}
%\end{equation}
%%
%Here $x(t) - X(t-1)$ serves as an independent correction
%from data point at time $t$ to the previous average value,
%and this correction is absorbed into the average
%with a weight of $1/t$,
%which is the inverse sample size.

%We can use Eq. \eqref{eq:avx_recur} to incrementally
%improve the estimated
%bias potential given by Eq. \eqref{eq:vcorr_equil}.
%
Assuming that we can linearize the histogram correction
in Eq. \eqref{eq:vcorr_equil} as
$\ln x \approx x - 1$ % (for $x \approx 1$),
and apply it to the instantaneous histogram, $h_i(t)$,
then the independent estimate of bias potential in each step
is given by
%
\begin{equation*}
  \hat u_i(t) \approx u_i(t) + \frac{ h_i(t) } { p_i } - 1
  .
  \label{eq:uhatt}
\end{equation*}
%
The time average
$\hat U_i(t) \equiv (1/t) \sum_{\tau=1}^t \hat u_i(\tau)$
is
\begin{align}
  \hat U_i(t)
  &\approx
  \frac{1}{t} \sum_{\tau=1}^t
  \left[
    u_i(\tau) + \frac{ h_i(\tau) } { p_i } - 1
  \right]
  .
  \label{eq:Uhatt}
\end{align}
%
For a fixed bias potential, the above formula
recovers the linearize Eq. \eqref{eq:vcorr_equil}.
%
But with the single-bin updating scheme,
Eq. \eqref{eq:wl_update}
we have, up to a constant shift,
%
\begin{align}
  \hat U_i(t)
  &\approx
  \frac{1}{t} \sum_{\tau=1}^t
  \left[
    u_i(\tau) +
    \frac{ u_i(\tau + 1) - u_i(\tau) } { \alpha(t) }
  \right]
  ,
  \notag
  \\
  &\approx
  \Delta u_i
  +
  \sum_{\tau=1}^t
  \left[
    1 + \frac{1}{\alpha(\tau - 1)} - \frac{1}{\alpha(\tau)}
  \right]
  \frac{ u_i(\tau) } {t}
  ,
  \label{eq:Uhat_sum}
\end{align}
where
$$
\Delta u_i \equiv \frac{u_i(t+1)}{t \, \alpha(t)}
- \frac{u_i(1)}{t \, \alpha(0)}.
$$
%
With the inverse-time schedule, Eq. \eqref{eq:alpha_invt},
each term in the square brackets vanishes,
and if we further define $\alpha(0) \to \infty$
$$
u_i(t+1) = \hat U_i(t),
$$
the bias potential represents
the runtime average of the correction, $\hat u_i(t)$.
%
Another special case is $\alpha(\tau) = 0$ for $\tau = 0, 1, \dots$,
then Eq. \eqref{eq:Uhat_sum} is reduced to a plain average
$$
\hat U_i(t) \approx \frac 1 t \sum_{\tau=1}^t u_i(\tau).
$$
where we have assumed an equilibrium condition
such that $u_i(t+1) \approx u_i(1)$.

For other updating schemes and schedules, we may still use
Eq. \eqref{eq:Uhatt} to construct a correction:
%
\begin{align}
  \hat U_i(t)
  &\approx
  U_i(t)
  +
  \frac{ H_i(t) } { p_i } - 1
  \approx
  U_i(t)
  +
  \ln \frac{ H_i(t) } { p_i }
  ,
  \label{eq:UHcorr}
\end{align}
%
where
\begin{align}
  U_i(t)
  &= \frac 1 t \sum_{\tau = 1}^t u_i(\tau)
  ,
  \notag
  %\label{eq:uav_def}
  \\
  H_i(t)
  &= \frac 1 t \sum_{\tau = 1}^t h_i(\tau)
  .
  \label{eq:hav_def}
\end{align}
%
%Thus, the bias potential from a WL simulation
%(which typically uses the single-bin updating scheme)
%under the inverse-time schedule
%would reach a similar precision
%as an ideal multicanonical simulation
%with fixed and nearly perfect bias potential.
%the precision at long times.
%%
%Thus, it is unnecessary to fix the bias potential
%after an adaptive FDS simulation has reached
%sufficiently convergent just for better precision
%of the final PMF.
